---
title: "StormData Analysis For Effects On Public Health & Economy"
author: "C.M. Oh"
date: "Friday, October 24, 2014"
output: html_document
---

# Synopsis
This document describes 'StormData'(of USA 1950~2011) analysis and its results.  
Most harmful weather events for public health & economy are calculated as the intersection of top-5 'mean set' and 'sum set'.  
In this respect, most harmful weather events for public health are "TORNADO" ,"(EXCESSIVE) HEAT" and "THUNDERSTORM WIND". Most harmful ones for economic properties are "HURRICANE","WINTER STORM" and "TROPICAL STORM", and ones for agricultural crop are "HURRICANE", "EXCESSIVE WETNESS", "COLD" AND "FROST".  
The "HEAT" event appears commonly in the public health areas( FATALITIES and INJURIES), and the "HURRICANE" in the economic damage areas(PROPERTIES and CROP).  

# Data Processing
### libraries && utility function
```{r, warning=FALSE}
Sys.setlocale(category="LC_ALL", locale="C")
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(R.utils))
```

```{r, echo=TRUE}
### Utility Functions
getFile <- function(fileUrl, destnm) {
    setInternet2(TRUE)
    if (!file.exists('data')) {
        dir.create('data')
    }
     if (file.exists(destnm)) {
         #message(cat(destnm,'already exists!'))
         #cat("------",list.files('./data'),"------",sep="\n")
         return(destnm)
     }
    download.file(fileUrl, destnm, mode="wb")
    return(destnm)
}
```

### download data file, preprocessing && load into data.table.
You need to install python and download preprocessing script from here:  
   https://raw.githubusercontent.com/Chang-Min/weather_analysis/master/preprocess.py
```{r,cache=TRUE}
data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
unzipped_file <- "data/StormDB-unzipped.csv"
data_file <- "data/StormDB.csv"  # preprocessed data file
if (!file.exists(unzipped_file)) {
  zipped_file <- paste(unzipped_file, "bz2", sep=".")
  getFile(data_url, zipped_file)
  bunzip2(zipped_file, unzipped_file)
  preproc_cmd = "python preprocess.py"
  system(preproc_cmd)
}
storm.db <- fread(data_file)
```

### event type normalizing function
```{r}
normalizeEvt <- function(nm) {
  if (length(grep('hurricane', nm, ignore.case=T))) {
    return("hurricane")
  } else if (length(grep('tornado',nm, ignore.case=T))) {
    return("tornado")
  } else if (length(grep('tstm|thunderstorm', nm, ignore.case=T))) {
    return("thunderstorm wind")
  } else if (length(grep('winter storm|winter weather',nm, ignore.case=T))) {
    return("winter storm")
  } else if (length(grep('cold|freeze|low temperature',nm, ignore.case=T))) {
    return("cold")
  } else if (length(grep('high seas|heavy sea|surf|high waves',nm,
                         ignore.case=T))) {
    return("high surf")
  } else if (length(grep('coastal flood',nm, ignore.case=T))) {
    return("coastal flood")
  } else if (length(grep('tropical storm',nm, ignore.case=T))) {
    return("tropical storm")
  } else if (length(grep('rain/snow',nm, ignore.case=T))) {
    return("rain/snow")
  } else if (length(grep('rain/wind',nm, ignore.case=T))) {
    return("rain/wind")
  } else if (length(grep('storm surge',nm, ignore.case=T))) {
    return("storm surge")
  } else if (length(grep('fire',nm, ignore.case=T))) {
    return("wild fire")
  } else if (length(grep('hail',nm, ignore.case=T))) {
    return("hail")
  } else if (length(grep('snow',nm, ignore.case=T))) {
    return("snow")
  } else if (length(grep('heat',nm, ignore.case=T))) {
    return("heat")
  } else if (length(grep('rain|rainfall',nm, ignore.case=T))) {
    return("rain")
  } else if (length(grep('slide',nm, ignore.case=T))) {
    return("landslide")
  } else if (length(grep('flood|high water|fld|high swell',nm, ignore.case=T))) {
    return("flood")
  } else if (length(grep('fog',nm, ignore.case=T))) {
    return("fog")
  } else if (length(grep('hypothermia|hyperthermia',nm, ignore.case=T))) {
    return("hyperthermia")
  } else if (length(grep('wind',nm, ignore.case=T))) {
    return("wind")
  } else if (length(grep('avalanc',nm, ignore.case=T))) {
    return("avalanche")
  } else if (length(grep('glaze|ice|icy',nm, ignore.case=T))) {
    return("glaze/ice storm")
  } else { return (nm) }
}
```
### make fatalities & injuries table by event type from storm database
```{r}
sum.ph.fat <- storm.db[,sum(FATALITIES),by=EVTYPE][V1>0]
sum.ph.inj <- storm.db[,sum(INJURIES),by=EVTYPE][V1>0]
mean.ph.fat <- storm.db[,mean(FATALITIES),by=EVTYPE][V1>0]
mean.ph.inj <- storm.db[,mean(INJURIES),by=EVTYPE][V1>0]
## normalize event type
sum.ph.fat <- sum.ph.fat[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
sum.ph.inj <- sum.ph.inj[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
mean.ph.fat <- mean.ph.fat[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
mean.ph.inj <- mean.ph.inj[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
norm.sum.ph.fat <- sum.ph.fat[,sum(V1), by=EVT]
norm.sum.ph.inj <- sum.ph.inj[,sum(V1), by=EVT]
# sum of means needed for normalized mean.
norm.mean.ph.fat <- mean.ph.fat[,sum(V1), by=EVT]
norm.mean.ph.inj <- mean.ph.inj[,sum(V1), by=EVT]
```



### make economic damage tables by event type from the database
```{r}
calcDamage <- function(dmg, exp) {
  if (exp == "B" || exp == "b") {
    return(dmg * 1000000000)
  } else
  if (exp == "M" || exp == "m") {
    return (dmg * 1000000)
  } else
  if (exp == "K" || exp == "k") {
    return (dmg * 1000)
  } else {
   return(dmg)
  }
}
sum.ed.pro <- storm.db[,sum(calcDamage(PROPDMG,PROPDMGEXP)), by=EVTYPE][V1>0]
sum.ed.cro <- storm.db[,sum(calcDamage(CROPDMG,CROPDMGEXP)), by=EVTYPE][V1>0]
mean.ed.pro <- storm.db[,mean(calcDamage(PROPDMG,PROPDMGEXP)), by=EVTYPE][V1>0]
mean.ed.cro <- storm.db[,mean(calcDamage(CROPDMG,CROPDMGEXP)), by=EVTYPE][V1>0]
## get normalized table
sum.ed.pro <- sum.ed.pro[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
sum.ed.cro <- sum.ed.cro[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
mean.ed.pro <- mean.ed.pro[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
mean.ed.cro <- mean.ed.cro[,EVT:=unlist(Map(normalizeEvt,EVTYPE))]
norm.sum.ed.pro <- sum.ed.pro[,sum(V1), by=EVT]
norm.sum.ed.cro <- sum.ed.cro[,sum(V1), by=EVT]
norm.mean.ed.pro <- mean.ed.pro[,sum(V1), by=EVT]
norm.mean.ed.cro <- mean.ed.cro[,sum(V1), by=EVT]
```


# Results

#### most harmful events for public health(fatalities)
```{r}
TOP5.SUM.PH.FAT <- head(norm.sum.ph.fat[order(V1, decreasing=T)],5)
TOP5.MEAN.PH.FAT <- head(norm.mean.ph.fat[order(V1, decreasing=T)],5)
intersect(TOP5.SUM.PH.FAT$EVT, TOP5.MEAN.PH.FAT$EVT)
```

#### most harmful events for public health(injuries)
```{r}
TOP5.SUM.PH.INJ <- head(norm.sum.ph.inj[order(V1, decreasing=T)],5)
TOP5.MEAN.PH.INJ <- head(norm.mean.ph.inj[order(V1, decreasing=T)],5)
intersect(TOP5.SUM.PH.INJ$EVT, TOP5.MEAN.PH.INJ$EVT)
```

#### most harmful events for economic damage(properties)
```{r}
TOP5.SUM.ED.PRO <- head(norm.sum.ed.pro[order(V1, decreasing=T)],5)
TOP5.MEAN.ED.PRO <- head(norm.mean.ed.pro[order(V1, decreasing=T)],5)
intersect(TOP5.SUM.ED.PRO$EVT, TOP5.MEAN.ED.PRO$EVT)
```

#### most harmful events for economic damage(agricultural crop)
```{r}
TOP5.SUM.ED.CRO <- head(norm.sum.ed.cro[order(V1, decreasing=T)],5)
TOP5.MEAN.ED.CRO <- head(norm.mean.ed.cro[order(V1, decreasing=T)],5)
intersect(TOP5.SUM.ED.CRO$EVT, TOP5.MEAN.ED.CRO$EVT)
```

#### damage graphs for top-5 events
```{r}
barplot(TOP5.MEAN.PH.FAT$V1, ylab="Mean Fatalities",col=rainbow(5),main="Top-5 mean fatalities for each weather events ")
legend("topright", TOP5.MEAN.PH.FAT$EVT, cex=1.2, fill=rainbow(5))
```
```{r}
barplot(TOP5.MEAN.ED.PRO$V1, ylab="Mean Properties Damage",col=rainbow(5),main="Top-5 mean property damage for each weather events ")
legend("topright", TOP5.MEAN.ED.PRO$EVT, cex=1.2, fill=rainbow(5))
```